Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-05-15 Ben_Manuscripts/hdphmm/Draft.tex
New references and review of hmms

  In their most basic form, hidden Markov models (HMMs) are used to generate
  a series of latent, or hidden, states dependent only on the previous state.~\cite{rabiner_tutorial_1989}
  Propagating an HMM generates sequential but independent observations from 
  the emission distributions associated with each hidden state. Transitions 
  between states are mathematically defined by a matrix of transition probabilities.
  Given a series of observations and a known number of hidden states, the 
  Baum-Welch algorithm is the most commonly used way to determine the hidden
  state sequence and the underlying parameters of a standard HMM.~\cite{baum_maximization_1970}
  It also possible to achieve the same results using a Bayesian 
  framework.~\cite{scott_bayesian_2002,jasra_markov_2005} One can place priors
  on the transition probabilities and parameters of the hidden states' emission
  distributions in order to iteratively infer their values and to estimate the hidden state
  sequence. HMMs have been used in a broad range of applications including speech
  recognition, weather forecasting and bioinformatics.~\cite{juang_on_1984,hughes_non-homogeneous_1999,yoon_hidden_2009}
  %BJC4: comment on usage in MD simulations. MSMs?

  By placing a hierarchical Dirichlet process (HDP) prior on the transition 
  probabilities, it is possible to extend HMMs to an unknown number of hidden
  states.~\cite{teh_hierarchical_2006}
  \begin{itemize}
    \item We will refer to this as the HDP-HMM, although in literature it is 
    often called the infinite hidden Markov model.~\cite{beal_infinite_2002}
    \item The HDP-HMM elegantly overcomes the main shortcoming of standard
    HMMs, the need to choose the number of states, making its use more 
    advantageous in a diversity of applications.
    \item

    diverse applicability and improvement upon the shortcoming 
    of standard HMMs
    \item Applications...speaker diarization, etc.
  \end{itemize}

  One issue with the original formulation of HDP-HMM, is that it does not 
  distinguish self-transitions and transitions between different states.
  \begin{itemize}
    \item Practically, this can result in rapid switching between states and
    inferred state sequence representations that do a poor job of describing
    the observations. 
    \item Fox et al. addressed this issue by adding a `sticky' parameter.~\cite{fox_sticky_2007}
    \item The sticky HDP-HMM is useful in situations where the user has
    prior knowledge that state persistence is likely.
    \item This is important for our work where we expect state persistence 
    due to trapping.
    \item Applications: Speaker diarization, SPT (Chris' diffusion papers)
  \end{itemize}

  The usefulness of the sticky HDP-HMM has been further extended to include 
  temporal dependencies common to many real datasets.
  \begin{itemize}
    \item Fox et al. showed how to incorporate Markov jump linear systems (MJLSs)
    into infinite state models.~\cite{fox_nonparameteric_2009}
    \item An MJLS models each dynamical mode as a linear dynamical process.
    \item There are two MJLS classes: switching linear dynamical systems (SLDSs) 
    and switching vector autoregressive (VAR) processes.
    \item Switching VAR processes are similar to SLDSs except that they do not
    incorporate measurement noise.
    \item Since MD simulations do not contain measurement noise, we will model
    describe the dynamical modes exhibited by solutes in our MD simulations as 
    VAR processes and we will refer to our model as an application of the 
    HDP-AR-HMM, maintaining a distinction from HDP-SLDS models which can
    be useful for experimental measurement where there is measurement noise.
  \end{itemize}

  To our knowledge, this is a novel application of the HDP-AR-HMM.
  \begin{itemize}
    \item We use the temporally-correlated parameters of the model in order
    to learn physical mechanisms and describe solute kinetics.
    \item We go beyond the typical application of IHMMs.
  \end{itemize}

  We aim to address two primary scientific questions with this work:
  First, we want to know whether it is possible to predict macroscopic transport
  properties. We use the VAR parameters and state transition probability matrix of
  the IHMM in order to generate stochastic trajectory realizations with behavior 
  similar to that exhibited by MD. One can project these realizations onto much 
  longer timescales with computational ease extending into timescales where 
  regular diffusive behavior again occurs.
  Second, we want to know if the IHMM algorithm can help researchers
  efficiently uncover underlying transport mechanisms which give rise to different 
  solute dynamical behavior.
  We use the parameters of the states identified by the IHMM in order to infer 
  dominant solute-membrane interactions and transport mechanisms. We cluster similar 
  state parameters in order to reduce the state space and to understand which segments 
  of the solute trajectories exhibit similar behavior. We support our mechanistic 
  hypotheses via quantitative comparison to various physical membrane properties and
  solute-membrane interactions.

>>>>

  In their most basic form, hidden Markov models (HMMs) are used to generate
  a series of latent, or hidden, states dependent only on the previous state.~\cite{rabiner_tutorial_1989}
  Propagating an HMM generates sequential but independent observations from 
  the emission distributions associated with each hidden state. Transitions 
  between states are mathematically defined by a matrix of transition probabilities.
  Given a series of observations and a known number of hidden states, the 
  Baum-Welch algorithm is the most commonly used way to determine the hidden
  state sequence and the underlying parameters of a standard HMM.~\cite{baum_maximization_1970}
  It also possible to achieve the same results using a Bayesian 
  framework.~\cite{scott_bayesian_2002,jasra_markov_2005} One can place priors
  on the transition probabilities and parameters of the hidden states' emission
  distributions in order to iteratively infer their values and to estimate the hidden state
  sequence. HMMs have been used in a broad range of applications including speech
  recognition, weather forecasting and bioinformatics.~\cite{juang_hidden_1984,hughes_non-homogeneous_1999,yoon_hidden_2009}
  %BJC4: comment on usage in MD simulations. MSMs?

  By placing a hierarchical Dirichlet process (HDP) prior on the transition 
  probabilities, it is possible to extend HMMs to an unknown number of hidden
  states.~\cite{teh_hierarchical_2006} We will refer to this as the HDP-HMM,
  although in literature it is often called the infinite hidden Markov 
  model.~\cite{beal_infinite_2002} The HDP-HMM elegantly overcomes the main
  shortcoming of standard HMMs, the need to choose the number of states 
  \textit{a priori}, making its use more advantageous in a diversity of 
  applications including single particle trajectory analysis and economic 
  data interpretation.~\cite{hines_analyzing_2015,shi_identifying_2016} 
  This approach allows the complexity of the data to drive the complexity
  of the model.

  One issue with the original formulation of the HDP-HMM, is that it does 
  not distinguish self-transitions and transitions between different states.
  Practically, this can result in rapid switching between states and inferred
  state sequence representations that do a poor job of describing the 
  observations. Fox et al. addressed this issue by adding a `sticky' 
  parameter.~\cite{fox_sticky_2007} The sticky HDP-HMM is useful in situations
  where the user has prior knowledge that state persistence is likely.
  For example, it is unlikely that a conversation between two people would 
  rapidly switch between speakers.~\cite{fox_sticky_2011} This is also 
  important for our work where we expect state persistence due to trapping.

  The usefulness of the sticky HDP-HMM has been further extended to include 
  temporal dependencies common to many real datasets. State dynamics and 
  relaxation into new states are often correlated.~\cite{calderon_data-driven_2014}
  Fox et al. showed formulated a method that accounts for this behavior by
  incorporating Markov jump linear systems (MJLSs) into infinite state 
  models.~\cite{fox_nonparametric_2009} An MJLS models each dynamical mode
  as a linear dynamical process. Two MJLS sub-classes are switching linear
  dynamical systems (SLDSs) and switching vector autoregressive (VAR) processes.
  Switching VAR processes are similar to SLDSs except that they do not
  incorporate measurement noise. Since MD simulations do not contain measurement
  noise, we will model the dynamical modes exhibited by MD solute trajectories 
  as VAR processes and we will refer to our model as an application of the 
  sticky HDP-AR-HMM in order to maintaining its the even more general sticky
  HDP-SLDS models.

  To our knowledge, this is a novel application and analysis of the sticky 
  HDP-AR-HMM. Typical applications center around determining the number of
  states, the state sequence and occasional analysis of the emission 
  distributions. The addition of temporal correlation to the inference 
  procedure has primarily been used to improve state sequence 
  identification.~\cite{calderon_inferring_2015,hamada_modeling_2016}
  Not only do we use the HDP-AR-HMM to segment MD time series into distinct
  dynamical modes, but we use temporally-correlated parameters of the model
  in order to learn physical mechanisms and describe solute kinetics. 
  %Further, we use the model parameters to generate new time series trajectories.

  %BJC4: enumerate?
  We aim to address two primary scientific questions with this work:
  First, we want to know whether it is possible to predict macroscopic transport
  properties. We use the VAR parameters and state transition probability matrix of
  the IHMM in order to generate stochastic trajectory realizations with behavior 
  similar to that exhibited by MD. One can project these realizations onto much 
  longer timescales with computational ease extending into timescales where 
  regular diffusive behavior again occurs.
  Second, we want to know if the IHMM algorithm can help researchers
  efficiently uncover underlying transport mechanisms which give rise to different 
  solute dynamical behavior.
  We use the parameters of the states identified by the IHMM in order to infer 
  dominant solute-membrane interactions and transport mechanisms. We cluster similar 
  state parameters in order to reduce the state space and to understand which segments 
  of the solute trajectories exhibit similar behavior. We support our mechanistic 
  hypotheses via quantitative comparison to various physical membrane properties and
  solute-membrane interactions.
